Extending K-Means Clustering with Ptolemy’s Inequality \thanks{This work is partly funded by DFG grant No. 454630593: EPIX – Efficient Ptolemaic Indexing.}

### 2025-01-14 paper/content/5_results.tex
put paper into its own subdir

\subsection{Experiments}
\todo{If we claim to have implemented the classic k-means, we also have to show that in our plots.}
Our variant of the k-Means algorithm is tested by implementing the classical k-Means algorithm (Lloyd's algorithm), Elkan's algorithm, and its novel, Ptolemy-based extension in Python.

To provide a fair comparison of the algorithms,
we measure the number of distance calculations instead of wall-clock time.
As distance calculations account for the majority of computing time in k-Means algorithms, they offer a clear measure of efficiency regardless of implementational optimizations.

@@

@@

\textbf{Limitations.}
The presented experiments are subject to three primary constraints:
(i) While synthetic random datasets provide valuable best-case scenarios, more extensive real-world data with appropriate grouping structures is necessary for comprehensive evaluation.
(ii) The significant reduction in distance calculations may not directly translate to improved execution times, as Ptolemy's bounds are computationally more expensive than the triangle inequality.
(iii) Most critically, the limited sample size and fixed initial conditions, particularly the lack of variation in cluster initializations, severely restrict the generalizability of individual measurements.

>>>>

\subsection{Experiments}
Our variant of the k-Means algorithm is tested by implementing the classical k-Means algorithm (Lloyd's algorithm), Elkan's algorithm, and its novel, Ptolemy-based extension in Python.

@@

To provide a fair comparison of the algorithms,
we measure the number of distance calculations instead of wall-clock time.
Distance calculations account for the majority of computing time in k-Means algorithms and do not depend on the implementation of the algorithms to machine code.
In comparison, execution time measurements are contingent on a number of interacting factors that render them difficult to reproduce.
These include the programming language and version, the amount and distribution of time spend on optimizing each algorithm's implementation, and the hardware architecture used to conduct the measurements.
Thus, it takes considerable effort and care to make time measurements objective, reproducible, and comparable.
Consequently, this short paper focuses on the theoretical, implementation-independent aspects of the discussed algorithms to determine whether a more extensive study is justified.

@@

\textbf{Limitations.}
The presented experiments are subject to three primary constraints:
(i) While synthetic random datasets provide valuable best-case scenarios, more extensive real-world data with appropriate grouping structures is necessary for a comprehensive evaluation.
(ii) The significant reduction in distance calculations may not directly translate to improved execution times, as Ptolemy's bounds are computationally more expensive than the triangle inequality.
(iii) Most critically, the limited sample size and fixed initial conditions, particularly the lack of variation in cluster initializations, severely restrict the generalizability of individual measurements.

### 2025-01-14 paper/content/6_conclusion.tex
put paper into its own subdir

% !TeX encoding = UTF-8
% !TeX program = pdflatex
% !BIB program = bibtex

\input{content/preamble}

\begin{document}
%%% Mehrere Autoren werden durch \and voneinander getrennt.
%%% Die Fußnote enthält die Adresse sowie eine E-Mail-Adresse.
%%% Das optionale Argument (sofern angegeben) wird für die Kopfzeile verwendet.
\title[Extending K-Means Clustering with Ptolemy’s Inequality]{Extending K-Means Clustering with Ptolemy’s Inequality
\thanks{This work is partly funded by DFG grant No. 454630593: EPIX – Efficient Ptolemaic Indexing.} 
}
%% \subtitle{Untertitel / Subtitle} % if needed
\author[1]{Max Pernklau}{max.pernklau@fernuni-hagen.de}{0009-0007-5520-4093}
\author[1]{Nikita Averitchev}{naveritchev@gmail.com}{}
\author[1]{Christian Beecks}{christian.beecks@fernuni-hagen.de}{0009-0000-9028-629X}
% \author[1]{Firstname4 Lastname4}{firstname4.lastname4@affiliation1.org}{0000-0000-0000-0000}%
\affil[1]{FernUniversität in Hagen\\Chair of Data Science\\ 58084 Hagen\\Germany}

\maketitle
\begin{abstract}
	Clustering is a fundamental data analytics operation in the field of unsupervised learning. Given a database of unknown structure, clustering aims to discover the inherent structure of the data objects according to similarity, such that similar data objects are grouped together, while dissimilar ones are separated in different groups or clusters.
	In this study, we focus on partitioning methods, specifically the \emph{k-means clustering} approach, which minimizes the intra-cluster variance.
	As the standard algorithmic approach to k-means clustering, the Lloyd algorithm, is neither efficient nor scalable, various adaptations and modifications have been developed, resulting in the family of fast k-means clustering algorithms.\\
    In this short paper, we extend the clustering algorithm proposed by Elkan with Ptolemy's inequality to prune superfluous distance calculations. It is not our intention to compete with the state-of-the-art algorithmic solutions for the k-means clustering problem; rather, we seek to investigate the potential of Ptolemy’s inequality to further enhancements in the standard algorithm’s performance, particularly in terms of reducing computations associated with distance evaluations.

	% neither tri → pto nor pto → tri is true
	% which generalizes the triangle inequality,

	%where the Lloyd algorithm is a prominent and de facto standard implementation.

	%The resulting groups, which are denoted as clusters, then represent the clustering structure of the database. A

\end{abstract}
\begin{keywords}
	k-means clustering \and Lloyd algorithm \and Ptolemy's inequality %Keyword1 \and Keyword2
\end{keywords}
%%% Beginn des Artikeltexts

%\todo{Americanize spelling (thesis uses chiefly British terms)}

\section{Introduction}

%intro
Clustering is a fundamental data-analytics operation in the domain of data science.
The basic task is to divide a set of data objects into different groups or clusters, such that objects within the same cluster have sufficient similarities;
meanwhile objects in different clusters should have significant differences.
The collection of clusters reflects the inherent structure of the underlying data-generating process and is denoted as a clustering.

The need for clustering arises in various scientific or economic application domains \cite{ezugwu2022comprehensive,oyewole2023data, gan2020data}, ranging from archaeology \cite{troiano2024comparative} and finance \cite{cai2016clustering} to industry \cite{lee2021technological} and zoology \cite{shen2021multivariate}, to name just a few. Especially in the field of unsupervised learning, clustering is utilized to extract structures from unlabeled data \cite{chander2023data}.

Alongside the diverse applications of clustering, unique domain-specific requirements and challenges arise, which are addressed by leveraging different families of clustering approaches \cite{xu2015comprehensive,han2012data}. These approaches, ranging from partitioning and hierarchical methods to density-based, grid-based and graph-based methods, are designed to accommodate varying data characteristics and domain-related requirements to facilitate efficient cluster analyses across diverse applications.

%different algorithms/problems
%kmeans/lloyd
%fast algorithms
%inefficient: distance calculations
%our proposal: ptolemaic
%our purpose

In this short paper, we focus on partitioning methods due to their simplicity in terms of interpretability and implementability \cite{DBLP:conf/iiwas/BeecksBHLSD22}. In this field, the \emph{k-means algorithm} \cite{bock2007clustering,hans2008origins,DBLP:journals/prl/Jain10,steinley2006k} has become one of the most influential clustering techniques \cite{DBLP:journals/kais/WuKQGYMMNLYZSHS08,olukanmi2019rethinking}. Although the term \emph{k-means algorithm} is explicitly credited to MacQueen \cite{macqueen1967}, its origins trace back to Steinhaus \cite{steinhaus1956division} in 1956, with its first application to data clustering by Forgy \cite{forgy1965cluster} in 1965. The widely recognized version in use today is the \emph{Lloyd algorithm} \cite{DBLP:journals/tit/Lloyd82}, introduced in 1982.

With the proliferation of complex data sources and the growing demand for more efficient clustering techniques, numerous \emph{fast k-means algorithms} have been introduced \cite{DBLP:conf/icml/Elkan03,DBLP:conf/sdm/Hamerly10,drake2012accelerated,hamerly2015accelerating,DBLP:conf/icml/NewlingF16,DBLP:conf/icml/DingZSMM15,DBLP:conf/sisap/SchubertLF21,DBLP:conf/sisap/LangS23}. These methods are designed to achieve good clustering results with significantly reduced computational cost. Moreover, most of them do not require any form of precomputation, making them directly applicable across a wide range of scenarios. The major objective of fast k-means algorithms is to reduce the number of distance evaluations required when assigning data objects to cluster centers and to safely prune cluster centers from the assignment process. For this purpose, distances are approximated via lower and upper bounds.

The Elkan algorithm \cite{DBLP:conf/icml/Elkan03} is a prominent representative of these algorithms and particularly suited for high-dimensional scenarios.
It maintains one upper bound and multiple lower bounds for each data object in order to apply different pruning criteria for each combination of data object and cluster center.
While the lower and upper bounds of Elkan's algorithm are directly derived from the triangle inequality, we propose to utilitze Ptolemy's inequality instead in order to increase the pruning performance.
With this extension, we do not aim to compete with the state of the art in k-means clustering;
rather, we intend to investigate the potential of Ptolemy’s inequality to further enhancements in the standard algorithm’s performance, specifically in terms of reducing computations associated with distance evaluations. We believe that our research findings offer a promising direction for future research and are beneficial for data scientists researchers and practitioners alike.

This short paper is structured as follows:
Section~\ref{Preliminaries} outlines the k-means problem and Elkan's approach of accelerating the algorithmic computation.
In Section~\ref{sec:contrib}, we show how Ptolemy's inequality can be applied to the aforementioned algorithm. %as our main contribution.
The results of our preliminary performance evaluation are detailed in Section~\ref{sec: results}, while Section~\ref{sec: conclusions} concludes this paper with a short outlook on future work.

%contribution

%structure

%\newpage

%\section{Related Work}
%\todo{remove this section if we run out of time}

\input{content/3_preliminaries}
\input{content/4_contribution}

\section{Preliminary Results}\label{sec: results}

\input{content/5_results}

\section{Conclusions and Future Work} \label{sec: conclusions}

@@

This short paper has addressed the problem of efficient data clustering by means of the k-means algorithm.
We have shown how to extend the more efficient clustering algorithm proposed by Elkan with novel distance approximations derived from Ptolemy's inequality.
The results of our preliminary performance evaluation indicate that our proposal achieves significant improvements in performance compared to Elkan's algorithm. %Particularly noteworthy is the increased efficiency with a large number of clusters.

Moreover, the use of the Ptolemy's inequality opens up new avenues for future work.
Besides conducting more extensive experiments to verify our findings,
it would be valuable to explore adaptations of this technique for other k-means algorithms and optimizations.
In particular for algorithms which rely on the triangle inequality to prune distance calculations, Ptolemy's inequality can potentially be leveraged to achieve an enhancement in performance, making these clustering algorithms more practical for a broader range of applications.

\bibliography{references,bibliography_thesis,bibliography_iiwas}
\end{document}

>>>>

\section{Conclusions and Future Work} \label{sec: conclusions}
This short paper addresses the problem of efficient data clustering using the k-means algorithm.
We show how the more efficient clustering algorithm proposed by Elkan can be extended with novel distance approximations derived from Ptolemy's inequality.
Our preliminary performance evaluation, which compares the number of total distance calculations needed to cluster a dataset, indicates that our extension achieves a significant improvement in performance compared to Elkan's algorithm.

We see two primary avenues for future work:
Firstly, more extensive experiments are required to verify our findings.
Specifically, execution times on a wider range of datasets must be measured to determine under which conditions the extension leads to real-world runtime savings.
Secondly, exploring adaptations of this approach to other k-means algorithms and optimizations could prove valuable.
In particular for algorithms which rely on the triangle inequality to prune distance calculations, Ptolemy's inequality might be leveraged to achieve better performance, making these clustering algorithms more practical for a broader range of applications.

%Particularly noteworthy is the increased efficiency with a large number of clusters.

@@

### 2025-01-14 paper/content/alg_elkan.tex
put paper into its own subdir

\section{Preliminaries}
\newcommand{\kmeans}{$k$-means problem\xspace}

\todo{explain structure in 3 sentences}

\subsection{The \kmeans}

To discuss the \kmeans, we first need to formalize the concept of a \emph{clustering algorithm}:
\begin{definition}[clustering algorithm]
	Given a target number of clusters $k$,
	let $F$ be a function that assigns each element from an input data set $D$ to a
	cluster index $i \in \{1, 2, \ldots, k\}$
	$$ F:D \to \{1, 2, \ldots, k\} \;.$$
\end{definition}
Notably, we restrict our analysis to clustering problems that use the $n$-dimensional Euclidean space as the domain of the input data set: $D \subset \mathbb{R}^n$.
Similarly, we assume that the similarity of the data points to be clustered is completely defined by the Euclidean distance $d(x,y)= || x-y ||_2 $.
The rationale behind this assumption will be made clear shortly in \autoref{pto}.

The heart of the \kmeans is the assignment of clusters so that the variance in similarity between all points in the same cluster should be reasonably small.
More formally:
\begin{definition}[\kmeans]
	Let $F$ be a clustering algorithm
	%$x \in D$ be an element from the input data set of a clustering algorithm $F$
	and let $C_i$ be the set of all elements assigned to cluster $i$:
	$$ C_i = \{x \in D \mid F(x) = i\} \,.$$
	Furthermore, let $\epsilon$ be the weighted sum of inter-cluster variances
	$$ \epsilon = \sum_{i=1}^k |C_i| \operatorname{Var}[C_i] = \sum_{i=1}^k \sum_{x_j \in C_i}  d^2(x_j, c_i)\,, $$
	where $c_i = \operatorname{E}[C_i]$ is the mean over all points in the cluster.

	Then, the map $F$ is called a solution to the \kmeans
	if $\epsilon$ cannot be made smaller by changing the assignment of any one element to a different cluster.
\end{definition}
Note that with this definition, the common $k$-means algorithms, like Lloyd's, provides solution to the \kmeans,
as we do not require that $F$ finds the smallest possible $\epsilon$,
just a local minimum of $\epsilon$.
This is consistent with the colloquial notion that solutions to the \kmeans ``can be found'',
even when an optimal solution (an NP-hard problem) is not practically obtainable \cite{}.

The most prevalent algorithm employed to identify valid solutions to the \kmeans is the Lloyd algorithm.
For convenience, it is reproduced in \autoref{alg:lloyd};
given its widespread use, we will only cover its most significant characteristics as they relate to our analysis.

\begin{algorithm}
	\caption{k-Means Algorithm}
	\label{alg:lloyd}

	\textbf{Input:} \( k \): Number of clusters, \( D \): Dataset containing \( n \) objects

	\textbf{Output:} Assignment of each $x_i \in D$ to a clusters

	\begin{algorithmic}[1]
		\State Initialize \( k \) cluster centers \( \{c_1, c_2, \dots, c_k\} \) arbitrarily from \( D \)
		\Repeat
		\label{algstep:assign}
		\State \textcolor{gray}{// Assign \( x_i \) to the closest cluster}
		\For{each object \( x_i \) in \( D \)}
		\State \( C_j \leftarrow C_j \cup \{x_i\} \) where \( j = \underset{j}{\arg\min} \; d^2(x_i, c_j) \)
		\EndFor
		\State \textcolor{gray}{// Update cluster center \( c_j \)}
		\For{each cluster \( C_j \)}
		\State \( c_j \leftarrow \frac{1}{|C_j|} \sum_{x_i \in C_j} x_i \)
		\EndFor
		\Until{no change in cluster assignments}
	\end{algorithmic}
\end{algorithm}

For each iteration of Lloyds algorithm, the assignment step (\autoref{algstep:assign}) needs to identify the closest cluster center for each element from the input dataset $D$.
To do so, Lloyd's algorithm calculates the distances between each cluster and each element from $D$, requiring $k\cdot|D|$ total distance evaluations per iteration.
These frequent distance evaluations can make up a significant share of the algorithms computational costs, especially for high-dimensional datasets.

\subsection{Accelerating $k$-means}

As we have seen, Lloyd's algorithm requires numerous evaluations of the distance function.
However, the exact distances do not actually need to be calculated explicitly for every element, as it is sufficient to identify which cluster center is the closest to a given element.
This is often possible from geometric considerations alone, when lower and upper distance bounds are available.
Two properties of the Euclidean space are of particular use here:
The triangle inequality
$$ d(x,y) \leq d(x,z) + d(z,y) $$
and Ptolemy's inequality
$$ d(x, y)\cdot d(v, u) \leq d(x, v) \cdot d(y,v) + d(x, u) \cdot d(y, v)$$
can provide lower and upper bound on the distance between two points given two (respectively five) other distances.

Numerous algorithms have been developed that exploit the former inequality to avoid explicit distance evaluations. However, to the best of our knowledge, the latter has yet to be employed for the purposes of accelerating k-means clustering.

Therefore, we define Elkan's algorithm \cite{}, which only uses the triangle inequality, here and extend it by incorporating Ptolemy's inequality in \ref{}.

\todo{insert Elkan's algo here? Or can we do it in any other way to save some space?}

\todo{explain Hetland's lower (upper) bound notation ($d^-$ and $d^+$)}
\begin{theorem}[Ptolemaic Bounds]
	\label{thm4.2}
	\todo{we might want to explain what the points are supposed to do before showing this theorem.
		Or, alternatively, we just present this as a purely mathematical statement with no preceding motivation.}
	Let \(a, b, c, c' \in \mathbb{R}^n\) and let \(d(\cdot, \cdot)\) be the Euclidean distance. The distance \(d(a, c')\) can be bounded as follows:
	\begin{gather}
		\frac{1}{d(b, c)} \cdot \max \left\{
		\begin{array}{l}
			d(a, b)^- \cdot d(c, c') - d(a, c)^+ \cdot d(b, c'), \\
			d(a, c)^- \cdot d(b, c') - d(a, b)^+ \cdot d(c, c')
		\end{array}
		\right\} \leq d(a, c')
	\end{gather}
	and
	\begin{gather}
		d(a, c') \leq \frac{1}{d(b, c)} \cdot \left( d(a, c) \cdot d(b, c') + d(a, b) \cdot d(c, c') \right)
	\end{gather}
\end{theorem}
\begin{proof}
	\todo{convert to passive voice}
	\todo{expand context to summarize all steps that lead to this proof}
	The lower bound corresponds to the derivation in equation \ref{eq4.4} and corresponds to the arrangement of the two pivot objects. Replace some distances with the respective lower or upper bound for the distance. The lower bounds can be used as they are always smaller than the exact distance. The upper bound must be used due to the subtraction. This means that the inequality still applies. The upper bound is obtained by using Ptolemy's inequality:

	\begin{equation*}
		\begin{aligned}
			\label{eq4.7}
			d(a,c') \cdot d(b,c) & \leq d(a, c) \cdot d(b, c') + d(a, b) \cdot d(c, c')                \\
			d(a,c')              & \leq \frac{d(a, c) \cdot d(b, c') + d(a, b) \cdot d(c, c')}{d(b,c)}
		\end{aligned}
	\end{equation*}
	By replacing some distances with upper boundaries you get the desired upper boundaries.
\end{proof}

\section{Applying Ptolemy's Inequality to the Elkan's Algorithm}

\subsection{Integration in k-Means}
Theorem \ref{thm4.2} will now be applied in a similar context to Elkan's algorithm \ref{alg:elkan}, with the goal of integrating these new bounds into steps 5 and 6 of the algorithm.

Consider the data set \(D = \{x_1, \ldots, x_n\}\). For each \(x_i\), there is an upper \(u_i\) and lower bounds \(l_{ij}\) as in Elkan's algorithm. The new cluster center assigned in a iteration is denoted by \(c_i^{\text{new}}\), and the old cluster center by \(c_i^{\text{old}}\). To use Ptolemy's Inequality, one additional point is needed. The idea is to use an even older cluster center \(c_i^{\text{old,old}}\) obtained two iterations ago.

Overall, to obtain a new upper bound for the distance between the datapoint \(x_i\) and the new cluster center \(c_j^{\text{new}}\), consider the old cluster center \(c_j^{\text{old}}\) from the last iteration to which \(x_i\) was assigned, and the even older cluster center \(c_j^{\text{old,old}}\) from two iterations ago. Utilize the old upper bounds that were valid for the respective iteration.

\begin{equation*}
	\begin{aligned}
		d(x_i,c_i^{old})      & \leq u_i^{old}      \\
		d(x_i,c_i^{old, old}) & \leq u_i^{old, old}
	\end{aligned}
\end{equation*}

Similarly, to obtain a new lower bound for the distance between the datapoint \(x_i\) and the new cluster center \(c_j^{\text{new}}\), the old lower bounds from the respective iterations are used:

\begin{equation*}
	\begin{aligned}
		l_{i,j}^{\text{old}}     & \leq d(x_i, c_j^{\text{old}})     \\
		l_{i,j}^{\text{old,old}} & \leq d(x_i, c_j^{\text{old,old}})
	\end{aligned}
\end{equation*}

Using Theorem \ref{thm4.2}, the following upper bound for the distance are obtained:
\begin{equation}
	\label{eq4.7}
	d(x_i, c_j^{\text{new}}) \leq u_i^{\text{new}} = \frac{1}{d(c_j^{\text{old}}, c_j^{\text{old,old}})} \cdot \left( u_i^{\text{old}} \cdot d(c_j^{\text{new}}, c_j^{\text{old,old}}) + u_i^{\text{old,old}} \cdot d(c_j^{\text{new}}, c_j^{\text{old}}) \right)
\end{equation}

and the new lower bound:
\begin{equation}
	\label{eq4.8}
	d(x_i, c_j^{\text{new}}) \geq l_{i,j}^{\text{new}} = \frac{1}{d(c_j^{\text{old}}, c_j^{\text{old,old}})} \cdot \max \left\{
	\begin{array}{l}
		l_{i,j}^{\text{old,old}} \cdot d(c_j^{\text{old}}, c_j^{\text{new}}) - u_i^{\text{old}} \cdot d(c_j^{\text{new}}, c_j^{\text{old,old}}), \\
		l_{i,j}^{\text{old}} \cdot d(c_j^{\text{new}}, c_j^{\text{old,old}}) - u_i^{\text{old,old}} \cdot d(c_j^{\text{old}}, c_j^{\text{new}})
	\end{array}
	\right\}
\end{equation}

You can see that the old upper boundaries are required to calculate the new lower boundaries.

Let us now look at how these bounds can be integrated into Elkan's algorithm. It is important to save not only the cluster centres and the bounds of the last iteration, but also those of the penultimate iteration.

At the beginning, all distances to all cluster centres are calculated in the first iteration. The upper bound is set by the distance to the nearest cluster centre, and the lower bounds are determined by the direct distances from the point to the respective cluster centres. An iteration is then performed as with Elkan, which means that bounds from two iterations are now available. This makes it possible to apply the new bounds from equations \ref{eq4.7} and \ref{eq4.8}. The algorithm then iterates, as in the original Elkan algorithm using updating with new update conditions, until the cluster centres converge or a specified number of iterations is reached. It is always important to check whether $d(c_j^{\text{old}}, c_j^{\text{old,old}}) \neq 0$. If this is the case, the update is carried out as for Elkan. The algorithm is displayed in Algorithm \ref{alg:ptolemy}.

\begin{algorithm}
	\caption{Elkan's Algorithm extended with Ptolemy's inequality}
	\label{alg:ptolemy}

	\begin{algorithmic}

@@

		\State \textbf{Step 0:} Compute one iteration of Elkan's algorithm. Save old upper and lower bounds of the last two iterations.
		\Repeat
		\State \textbf{Step 1:}  Compute distances $d(c_i, c_j)$ between all centers, and calculate $s(c_i) = \frac{1}{2} \min_{c_j \neq c_i} d(c_i, c_j)$ for each center $c_i$.
		\Statex
		\State \textbf{Step 2:}  Retain points $x_i$ in their current clusters if $u(x_i) \leq s(c(x_i))$.
		\Statex
		\State \textbf{Step 3:}  For remaining points, consider $x_i$ for reassignment if:
		\begin{itemize}
			\item $c_j \neq c(x_i)$,
			\item $u(x_i) > l(x_i, c_j)$, and
			\item $u(x_i) > \frac{1}{2}d(c(x_i), c_j)$.
		\end{itemize}
		\Statex
		\State \textbf{Step 3a:}  If $r(x_i)$ is true, compute $d(x_i, c(x_i))$. Set $r(x_i) = \text{false}$. Otherwise, $u(x_i) =d(x_i, c(x_i))$
		\Statex
		\State \textbf{Step 3b:} If $d(x_i, c(x_i)) > l(x_i, c_j)$ or $d(x_i, c(x_i)) > \frac{1}{2}d(c(x_i), c_j)$, compute $d(x_i, c_j)$. Reassign $x_i$ to $C_j$ if $d(x_i, c_j) < d(x_i, c(x_i))$.
		\Statex
		\State \textbf{Step 4:} Compute the cluster centers as the centroids of the corresponding clusters $c'_j$.
		\Statex
		\State \textbf{Step 5:} If distance between two older cluster is not 0. Update lower bounds by using equation \ref{eq4.8}. Otherwise, update like in Elkan. Save the old lower bounds of the two last iterations.
		\Statex
		\State \textbf{Step 6:} If distance between two older cluster is not 0. Update upper bounds by using equation \ref{eq4.7}. Otherwise, update like in Elkan. Reset $r(x_i) = \text{true}$. Save old cluster and upper bounds of the last two iterations.
		\Statex
		\State \textbf{Step 7:} Replace each center $c_j$ with $c'_j$.

>>>>

\todo{original Elkan as presented in NiAv's bachelor thesis. has slightly excessive spacing}
\begin{algorithm}
	\caption{Accelerated k-Means Algorithm (Elkan)}
	\label{alg:elkan}
	\begin{algorithmic}

@@

		\Repeat
		\State \textbf{Step 1:}  Compute distances $d(c_i, c_j)$ between all centers, and calculate $s(c_i) = \frac{1}{2} \min_{c_j \neq c_i} d(c_i, c_j)$ for each center $c_i$.
		\Statex
		\State \textbf{Step 2:}  Retain points $x_i$ in their current clusters if $u(x_i) \leq s(c(x_i))$.
		\Statex
		\State \textbf{Step 3:}  For remaining points, consider $x_i$ for reassignment if:
		\begin{itemize}
			\item $c_j \neq c(x_i)$,
			\item $u(x_i) > l(x_i, c_j)$, and
			\item $u(x_i) > \frac{1}{2}d(c(x_i), c_j)$.
		\end{itemize}
		\Statex
		\State \textbf{Step 3a:}  If $r(x_i)$ is true, compute $d(x_i, c(x_i))$. Set $r(x_i) = \text{false}$. Otherwise, $u(x_i) =d(x_i, c(x_i))$
		\Statex
		\State \textbf{Step 3b:} If $d(x_i, c(x_i)) > l(x_i, c_j)$ or $d(x_i, c(x_i)) > \frac{1}{2}d(c(x_i), c_j)$, compute $d(x_i, c_j)$. Reassign $x_i$ to $C_j$ if $d(x_i, c_j) < d(x_i, c(x_i))$.
		\Statex
		\State \textbf{Step 4:} Compute the cluster centers as the centroids of the corresponding clusters $c'_j$.
		\Statex
		\State \textbf{Step 5:} Update lower bounds $l(x_i, c_j)$ for each $x_i$ and $c_j$ using $l(x_i, c_j) = \max \{ l(x_i, c_j) - d(c'_j, c_j), 0 \}$.
		\Statex
		\State \textbf{Step 6:} Update upper bounds $u(x_i)$ for each $x_i$ using $u(x_i) = u(x_i) + d(c'(x_i), c(x_i))$ and reset $r(x_i) = \text{true}$.
		\Statex
		\State \textbf{Step 7:} Replace each center $c_j$ with $c'_j$.
